---
title: "ASSIGNMENT02"
author: "Dibanisa Fakude"
date: "2024-03-03"
output:
  html_document:
    df_print: paged
---

```{r,echo=FALSE}

library(tidyverse)
library(lubridate)
library(stringr)
library(tidyverse)
library(lubridate)

# Read earthquake data and convert "Date" column to Date format
earthquake_data <- read.csv("~/EDA_ASSIGNMENTS/ASSIGNMENT2/Earthquakes 1965 - 2016.csv") %>%
  as_tibble() %>%
  mutate(Date_Time = mdy_hms(paste(Date, Time))) %>%
  select(-Date,-Time)

# Read query data and separate "Time" column into "Date" and "Time"
query_data <- read.csv("~/EDA_ASSIGNMENTS/ASSIGNMENT2/query.csv") %>%
  as_tibble() 



# Rename variables in query_data to match earthquake_data
query_data <- query_data %>%
  rename(Latitude = latitude, Longitude = longitude, Depth = depth, Magnitude = mag, Magnitude.Type = magType,
         ID = id, Type = type, Status = status, Location.Source = locationSource, Magnitude.Error = magError,
         Magnitude.Source = magSource, Azimuthal.Gap = gap, Root.Mean.Square = rms, Horizontal.Error = horizontalError,
         Depth.Error = depthError, Magnitude.Seismic.Stations = magNst, Date_Time = time)



# mKA THE FIRST LETTER UPPERCASES
query_data <- query_data %>%
  mutate(Type = str_to_title(Type))
#Changing the Date Time Column in query data to Date format
  query_data <- query_data %>%
  mutate(Date_Time = ymd_hms(Date_Time))

# Merge the datasets
merged_data <- bind_rows(earthquake_data, query_data)


```
Explore the range of magnitudes (Richter scale) in both datasets. Is there a reason to
exclude any measurements in either? Implement and discuss briefly.

```{r,echo=FALSE}


  
# Calculate the range of magnitudes in the first dataset
min_magnitude_df1 <- min(earthquake_data$Magnitude)
max_magnitude_df1 <- max(earthquake_data$Magnitude)

# Calculate the range of magnitudes in the second dataset
min_magnitude_df2 <- min(query_data$Magnitude)
max_magnitude_df2 <- max(query_data$Magnitude)

# Print the range of magnitudes in both datasets
cat("Range of magnitudes in the first dataset:", min_magnitude_df1, "-", max_magnitude_df1, "\n")
cat("Range of magnitudes in the second dataset:", min_magnitude_df2, "-", max_magnitude_df2, "\n")


# Create a dataframe with dataset names and their respective range of magnitudes
data_ranges <- data.frame(
  Dataset = c("Earthquake Data ", "Query Data"),
  Min_Magnitude = c(5.5, 5),
  Max_Magnitude = c(9.1, 8.2)
)

# Plot the ranges as horizontal lines
ggplot(data_ranges, aes(y = Dataset, x = Max_Magnitude, xmin = Min_Magnitude, xmax = Max_Magnitude, color = Dataset)) +
  geom_point(size = 3) +
  geom_errorbarh(height = 0.2) +
  labs(title = "Range of Magnitudes in Datasets",
       x = "Magnitude",
       y = "") +
  theme(plot.title = element_text(hjust = 0.5)+
  theme_minimal())
```

Report on the extent of missing data in the common variables. Is it worth keeping any
of the information for variables with missing data? Likewise for common variables with
a discrepancy in values. Implement and discuss briefly.
```{r,echo=FALSE}

library(dplyr)


# Define the variable names
variable_names <- c(
  "Latitude", "Longitude", "Type", "Depth", "Depth.Error",
  "Depth.Seismic.Stations", "Magnitude", "Magnitude.Type", 
  "Magnitude.Error", "Magnitude.Seismic.Stations", "Azimuthal.Gap",
  "Horizontal.Distance", "Horizontal.Error", "Root.Mean.Square",
  "ID", "Source", "Location.Source", "Magnitude.Source",
  "Status", "Date_Time", "nst", "dmin", "net", "updated", "place"
)

# Create a report on missing data for each variable
missing_report <- lapply(variable_names, function(var) {
  missing_count <- sum(is.na(merged_data[[var]]))
  data.frame(
    Variable = var,
    Missing_Count = missing_count
  )
})

# Combine the report into a single dataframe
missing_report_df <- do.call(rbind, missing_report)

# Print the report
print(missing_report_df)
#First Exxclude the ones with zeros
# Exclude variables with zero missing counts and arrange in descending order
missing_report_filtered <- missing_report_df %>%
  filter(Missing_Count != 0) %>%
  arrange(desc(Missing_Count))

# Plot bar graph
ggplot(missing_report_filtered, aes(x = reorder(Variable, -Missing_Count), y = Missing_Count)) +
  geom_bar(stat = "identity", fill = "black", alpha =0.6) +
  labs(title = "Missing Data Counts for Variables with Missing Values",
       x = "Variable", y = "Missing Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))  # Center the plot title
```


 Provide a summary of the types of events and then filter the observations to only include
```{r,echo=FALSE}

Summaryevents<-merged_data %>% 
  group_by(Type) %>% 
  summarise(mean = mean(Magnitude),
            maxMagnitude= max(Magnitude),
            minMagnitude = min(Magnitude),
            NumberofEvents = n())

Summaryevents

# Filter observations to include only "Earthquake" type events
earthquakes <- merged_data  %>%
  filter(Type == "Earthquake")
  
```

Add a categorical variable “Scale”, which classifies the earthquakes according to their
magnitude. Classify and name the categories according to the “Description” column in
the table on this page.
```{r,echo=FALSE}
earthquakes<- earthquakes %>% 
  mutate(Scale = case_when(
    Magnitude >= 1.0 & Magnitude <= 1.9 ~ "Micro",
    Magnitude >= 2.0 & Magnitude <= 2.9 ~ "Minor",
    Magnitude >= 3.0 & Magnitude <= 3.9 ~ "Slight",
    Magnitude >= 4.0 & Magnitude <= 4.9 ~ "Light",
    Magnitude >= 5.0 & Magnitude <= 5.9 ~ "Moderate",
    Magnitude >= 6.0 & Magnitude <= 6.9 ~ "Strong",
    Magnitude >= 7.0 & Magnitude <= 7.9 ~ "Major",
    Magnitude >= 8.0 & Magnitude <= 8.9 ~ "Great",
    Magnitude >= 9.0 & Magnitude <= 9.9 ~ "Extreme",
    TRUE ~ "Unknown"
  ))
  
```
The distributions of the features. Decide which features to investigate and provide the
appropriate plots/measurements.

```{r,echo=FALSE}
library(gridExtra)

# Select features for investigation
features_to_explore <- c("Magnitude", "Depth")

# Generate histograms for selected features
histograms <- lapply(features_to_explore, function(feature) {
  ggplot(earthquake_data, aes(x = !!sym(feature))) +
    geom_histogram(fill = "grey", color = "black", bins = 20) +
    labs(title = paste("Histogram of", feature),
         x = feature,
         y = "Frequency") +
    theme_bw()+
    theme(plot.title = element_text(hjust = 0.5))
})

# Combine histograms into one plot
combined_plot <- do.call(grid.arrange, histograms)

# Print combined plot
print(combined_plot)



```
The largest earthquakes, according to magnitude. Choose a number of them and present
and discuss in an appropriate way.

```{r,echo=FALSE}
# Filter earthquakes based on scale
large_earth <- earthquakes %>% 
  filter(Scale == "Extreme")

# Create a new dataframe containing latitude, longitude, magnitude, and scale information
largest_earthquakes <- data.frame(
  Latitude = large_earth$Latitude,
  Longitude = large_earth$Longitude,
  Magnitude = large_earth$Magnitude,
  Scale = large_earth$Scale
)

# View the resulting dataframe
largest_earthquakes

```
The relationship between depth and magnitude, is there any pattern worth noting? Note
that the Richter scale is logarithmic, such that an increase of 1 unit corresponds to a
tenfold increase in actual magnitude, and a 31-fold increase in energy. How does the
picture change when adjusting the magnitude accordingly?

```{r,echo=FALSE}
#The relationship between depth and Magnitude
cor(earthquakes$Depth,earthquakes$Magnitude)


# Scatter plot of depth versus magnitude
ggplot(earthquakes, aes(x = Depth, y = Magnitude)) +
  geom_point(alpha =0.5) +
  labs(title = "Depth vs Magnitude",
       x = "Depth",
       y = "Magnitude") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))

# Adjust magnitude values using logarithmic scale of Richter scale
earthquakes$log_Magnitude <- log10(earthquakes$Magnitude)

# Scatter plot of depth versus adjusted magnitude
ggplot(earthquakes, aes(x = Depth, y = log_Magnitude)) +
  geom_point(alpha =0.5) +
  labs(title = "Depth vs Adjusted Magnitude (log scale)",
       x = "Depth",
       y = "Adjusted Magnitude") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```
Choose one of the types of events that were removed and briefly investigate these events.
When and where did they take place? Provide a map showing this.

```{r,echo=FALSE}
# Load world map data into a dataframe named world_map
world_map <- map_data("world")

# Filter map data for Turkey
map_hawaii <- world_map %>% filter(region == "Hawaii")

# Filter volcanic eruption data and removing the deviated point
volcanic_eruptions <- merged_data %>%
  filter(Type == "Volcanic Eruption" & !ID == "us7000gc8r") 


# Plot the volcanic eruptions on a map
hawaii<-ggplot() +
  borders("world", colour = "gray50", fill = "gray80") +
    geom_point(data = volcanic_eruptions, aes(x = Longitude, y = Latitude), color = "red", size = 3) +
  theme_bw() +
  labs(title = "Locations of Volcanic Eruptions",
       x = "Longitude", y = "Latitude") +
  xlim(-160, -150) +
  ylim(19, 23)

city_labels <- data.frame(
  city = "Hawaii",
  lon = -155.844437,
  lat = 19.741755
)

Label <- hawaii +
  geom_text(data = city_labels, aes(x = lon, y = lat, label = city), size = 3, hjust = 0, vjust = -0.5)

Label

```
How do the frequency of earthquakes in each category compare to estimated frequencies,
like those given here or here?
Do we observe any patterns over time? Is this a definite function of time, or perhaps an
artefact of the data?

```{r,echo=FALSE}
#Temporal Investigation
#Summary 3
library(tidyr)
library(ggplot2)
earthquakes %>% 
  group_by(Scale) %>% 
  summarise(count=n(),
            averageMagnitude = mean(Magnitude))

# Calculate the frequency of earthquakes by scale and year
freq_earth_wide <- earthquakes %>%
  filter(Scale != "Unknown") %>%  # Filter out rows with "Unknown" in Scale column
  group_by(Scale, year = lubridate::year(Date_Time)) %>%
  count()

# Define colors for each earthquake scale
colors <- c("red", "green", "blue", "orange", "purple", "yellow")

# Create a line plot of frequency of earthquakes by scale and year
freq_earth_wide %>%
  ggplot(aes(x = year, y = n, color = Scale)) +
  geom_line() +  # Add lines
  facet_grid(~Scale) +  # Facet the plot by Scale
  labs(title = "Frequency of Earthquakes by Scale and Year",  # Add title and axis labels
       y = "Frequency",
       x = "Year") +
  scale_color_manual(values = colors) +  # Set colors for each scale
  theme_bw() +  # Use a black and white theme
  theme(
    plot.title = element_text(hjust = 0.5)  # Center the title horizontally
  )

```

4. Spatial Exploration
The country of T¨urkiye (Turkey). On 6 February 2023, a magnitude 7.8 earthquake
devastated the southern region. How does this compare with previous occurrences in
the region? How often do they experiences major or great earthquakes?

```{r,echo=FALSE}
#number 1
library(maps)

# Load world map data into a dataframe named world_map
world_map <- map_data("world")
mapturkey <- world_map %>% filter(region == "Turkey")
turkeydata1 <- earthquakes %>% filter(Date_Time <= as.Date("2023-02-06"))
title <- "Previous Earthquakes in The Turkey Region"
#Findinf the mean of the prevois earthquakes to compare with the currecnt 7.8
turkeydata1 %>% 
  group_by(Scale) %>% 
  summarise(MeanEarth= mean(Magnitude))
# Plotting
ear <- ggplot() +
  geom_map(data = mapturkey, map = mapturkey, 
           aes(x = long, y = lat, map_id = region), fill = "grey", color = "black", size = 0.5) +
  geom_point(data = turkeydata1, aes(x = Longitude, y = Latitude, color = Magnitude), size = 5) +
  scale_colour_gradient(low = "blue", high = "red") +
  ggtitle(title) +
  xlim(26.0, 45.0) +
  ylim(36.0, 42.0)+
theme_bw()

city_labels <- data.frame(
  city = c("Ankara", "Istanbul", "Izmir", "Antalya", "Adana"),
  lon = c(32.8541, 28.9784, 27.1428, 30.7133, 35.3228),
  lat = c(39.9356, 41.0082, 38.4192, 36.8969, 37.0015)
)

ear1 <- ear +
  geom_text(data = city_labels, aes(x = lon, y = lat, label = city), size = 3, hjust = 0, vjust = -0.5)

ear1

View(turkeydata1)
```


```{r,echo=FALSE}
#How often they experience major or great

# Load world map data into a dataframe named world_map
world_map <- map_data("world")
mapturkey <- world_map %>% filter(region == "Turkey")
turkeydata1 <- earthquakes %>% filter(Scale == "Major"| Scale == "Great")
title <- "earthquakes"

count(turkeydata1)



# Plotting
ear <- ggplot() +
  geom_map(data = mapturkey, map = mapturkey, 
           aes(x = long, y = lat, map_id = region), fill = "grey", color = "black", size = 0.5) +
  geom_point(data = turkeydata1, aes(x = Longitude, y = Latitude, color = Magnitude), size = 5) +
  scale_colour_gradient(low = "blue", high = "red") +
  ggtitle(title) +
  xlim(26.0, 45.0) +
  ylim(36.0, 42.0)+
theme_bw()

city_labels <- data.frame(
  city = c("Ankara", "Istanbul", "Izmir", "Antalya", "Adana"),
  lon = c(32.8541, 28.9784, 27.1428, 30.7133, 35.3228),
  lat = c(39.9356, 41.0082, 38.4192, 36.8969, 37.0015)
)

ear1 <- ear +
  geom_text(data = city_labels, aes(x = lon, y = lat, label = city), size = 3, hjust = 0, vjust = -0.5)

ear1

```

Southern Africa, including some area off the South African coast. How many earthquakes
have we experienced since 1965? Which was the largest?
```{r,echo=FALSE}

library(maps)
library(ggplot2)
library(dplyr)

# Filter map data for Southern Africa
world_map <- map_data("world")
map_southern_africa <- world_map %>%
  filter(region %in% c("South Africa", "Lesotho", "Eswatini", "Namibia", "Botswana", "Zimbabwe", "Mozambique"))

# Filter earthquakes data for Southern Africa
southern_africa_quakes <- earthquakes %>%
  filter(Latitude >= -35 & Latitude <= -10,
         Longitude >= 10 & Longitude <= 40,
         Date_Time >= as.Date("1965-01-01"))
count(southern_africa_quakes)
# Find the largest earthquake in Southern Africa
largest_earthquake <- southern_africa_quakes %>%
  filter(Magnitude == max(Magnitude))

# Print the number of earthquakes in Southern Africa since 1965 and the details of the largest earthquake
cat("Number of earthquakes in Southern Africa since 1965:", nrow(southern_africa_quakes), "\n")
cat("Details of the largest earthquake in Southern Africa since 1965:\n")
print(largest_earthquake)

# Plotting
ggplot() +
  geom_map(data = map_southern_africa, map = map_southern_africa, 
           aes(x = long, y = lat, map_id = region), fill = "grey", color = "black", size = 0.5) +
  geom_point(data = southern_africa_quakes, aes(x = Longitude, y = Latitude, color = Magnitude), size = 5) +
  scale_colour_gradient(low = "blue", high = "red") +
  ggtitle("Earthquakes in Southern Africa since 1965") +
  theme_bw()


```
An area/country of your choosing. This could be based on your global map, or because
you have an interest in a specific region for whatever reason.

```{r,echo=FALSE}
#Number 3
library(ggplot2)
library(maps)
library(dplyr)


# Filter map data for Japan
world_map <- map_data("world")
map_japan <- world_map %>%
  filter(region == "Japan")

# Filter earthquakes data for Japan since 1965
japan_quakes <- earthquakes %>%
  filter(Latitude >= 24 & Latitude <= 45,
         Longitude >= 122 & Longitude <= 153,
         Date_Time >= as.Date("1965-01-01"))

# Count the number of earthquakes in Japan since 1965
num_quakes_japan <- nrow(japan_quakes)

# Find the largest earthquake in Japan since 1965
largest_earthquake_japan <- japan_quakes %>%
  filter(Magnitude == max(Magnitude))

# Print the number of earthquakes in Japan since 1965 and the details of the largest earthquake
cat("Number of earthquakes in Japan since 1965:", num_quakes_japan, "\n")
cat("Details of the largest earthquake in Japan since 1965:\n")
print(largest_earthquake_japan)


```



